Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-02-13 response.tex

We have included CRBs for figures in the simulations section.  The following paragraph has been included:

``Figures~\zref{fig:plotphaseBPSK} and~\zref{fig:plotphaseQPSK} display the non-data aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996}.  To the authors' knowledge, simple expression for the CRB for $8$-PSK do not exist in the literature and so the CRB is only shown when $M=2$ (BPSK) and $M=4$ (QPSK).  These CRBs are derived for the non-data aided setting and are only valid for estimators that do not make use of pilots symbols, i.e., $\abs{P}=\emptyset$.''

@@

``This appears to be the first time that the statistical properties of a non-data-aided estimator for carrier amplitude have been analysed theoretically.''

and in the new introduction we affirm that

``while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.''

Thus, the techniques in our paper can potentially be deployed to accurately model the bias and variance of a non-data-aided SNR estimator in order to implement the reviewer's proposal.  We confirm, however, that analysing and implementing the reviewer's suggested technique is beyond the scope of the present paper which places primary interest in analysing the least squares estimator.

@@

\item\textbf{Comment:} \label{rev2:biasmse}
	By detailed inspection of Fig.\ 4 showing the jitter variance for amplitude estimates, it is easily observed that $|P| < L$ performs better than $|P| = L$ in the medium-to-low SNR range. In my mind this is somewhat strange, since it should be the other way round (similar to the evolution of the phase curves in Figs.\ 1Ð3 and 5). Can you comment on this phenomenon ?
	\\\textbf{Response:}
In the original manuscript we plotted the \emph{variance} of the amplitude estimator, not the mean square error. That is, the solid lines displayed the asymptotic variance of $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$ and each point from the Monte-Carlo simulations was computed as the unbiased error $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  We chose this approach in order to highlight the correctness of our asymptotic theory. This was described in the 4th paragraph of the old Section~8 containing the simulations section.

We do agree with the reviewer that this plot by itself initially appears misleading.  To alleviate this confusion in the new manuscript we instead plot the average square error computed in the usual way by Monte-Carlo simulation and the solid line is given by the asymptotic squared bias plus variance.  The following paragraph has also been included in the simulations section (Section~\zref{sec:simulations}) in the new manuscript.

@@

\end{enumerate}
\end{enumerate}

\section*{Reviewer 3}

\begin{enumerate}
\item
	\textbf{Comment:} 
The paper studies the estimation of amplitude and phase due to the distortion introduced by the channel in communication systems with $M$-PSK signals. The estimator is derived from the observation of $L$ noisy samples taken at a rate of one sample per symbol. Part of the $L$ observed samples corresponds to pilot transmitted symbols, that are already known a priori by the receiver, while the rest of samples are associated with data symbols which are unknown at the receiver. The focus of the article is to determine the amplitude and phase without considering the estimation of the transmitted data like this problem is usually treated in the literature. The work modifies the Mackenthun algorithm to include samples corresponding to pilot symbols while the Mackenthun paper calculates a least squares estimator for a similar problem but without pilot symbols.

The major contribution of this paper is the description of the asymptotic properties of the least squares estimator for the phase and amplitude. Although it seems technically correct, the authors do not make clear the relevance of their work to signal processing applications and to timely problems. They limit themselves in presenting several lemmas and proofs. Therefore, I cannot recommend this manuscript for publication in the IEEE Transactions on Signal Processing.
	\\\textbf{Response:}
We have greatly expanded the introduction of the manuscript to make the practical and theoretical importance of our results clear.  Also see our response to comment~\ref{rev4:shouldbettermotivate} of reviewer~4.
\end{enumerate}

@@

%	\comment{Stress practical relevance by showing ASRP PER results. In particular ST1, where the estimators are used to seed turbo sync. Mention trials and patents\dots}

%	\comment{Not clear what the reviewer means by ``existing uncoherent methods that do not require estimation of the amplitudes''.}

\item
Some additional comments:
\begin{enumerate}
\item\textbf{Comment:} \label{rev4:shouldbettermotivate}
	The authors should better motivate the interest in using their proposed method in actual communication systems. The lack of motivation for addressing the problem raised in the manuscript is a weak aspect of this work.
	\\\textbf{Response:}
We have expanded the introduction of the paper to better motivate the work.  In particular, the following paragraph has been added:

``That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\zref{thm:normality} in Section~\zref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing system parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.''

### 2014-02-13 paper.tex
added some comments at end re cleaning up the wild formula for 8psk computation
added Andre's comments regarding `hard decision' notation

To see this, let $a = \rho e^{j\theta}$ where $\rho > 0$ is real.  Now $LS$ from~\eqref{eq:LS} is to
\begin{align}
LS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:LSallparams}
\end{align}
We have slightly abused notation here by reusing $LS$. This should not cause confusion as $LS(a, \{d_i, i \in D\})$, $LS(\rho, \theta, \{d_i, i \in D\})$, and $LS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the sample MSE of the non-data-aided phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires selection of a function $F$ that transforms the amplitude of each sample prior to the final estimation step.  % (see equation (2) in~\cite{ViterbiViterbi_phase_est_1983}).  
Viterbi and Viterbi propose several viable alternatives, from which we have chosen $F(x) = 1$.  The Viterbi and Viterbi estimator is only applicable in the non-data-aided setting, i.e. when $\abs{P} = 0$.  The sample MSE of the least squares estimator (when $\abs{P} = 0$) and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate for some values of SNR.

Figures~\ref{fig:plotphaseBPSK} and~\ref{fig:plotphaseQPSK} display the non-data aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996}.  To the authors' knowledge, simple expression for the CRB for $8$-PSK do not exist in the literature and so the CRB is only shown when $M=2$ (BPSK) and $M=4$ (QPSK).  These CRBs are derived for the non-data aided setting and are only valid for estimators that do not make use of pilots symbols, i.e., $\abs{P}=0$.

@@

>>>>

To see this, let $a = \rho e^{j\theta}$ where $\rho > 0$ is real.  Now $LS$ from~\eqref{eq:LS} is to
\begin{align}
LS(\rho, \theta, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - \rho e^{j\theta} s_i }^2  \nonumber \\
&= \sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} d_i }^2. \label{eq:LSallparams}
\end{align}
We have slightly abused notation here by reusing $LS$. This should not cause confusion as $LS(a, \{d_i, i \in D\})$, $LS(\rho, \theta, \{d_i, i \in D\})$, and $LS(\{d_i, i \in D\})$ are easily told apart by their arguments.  For given $\theta$, the least squares estimator of the $i$th data symbol $d_i$ is given by minimising $\abs{ y_i - \rho e^{j\theta} d_i }^2$, that is,
\begin{equation}\label{eq:hatdfinxtheta}
\hat{d}_i(\theta) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\end{equation}
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  A word of caution, the notation $\round{\cdot}$ is often used to denote rounding to the nearest \emph{integer}.  This is not the case here.  If the function $\operatorname{round}(\cdot)$ takes its argument to the nearest integer then,
\[
\round{x} = \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right).
\] 
It is common to call $\hat{d}_i(\theta)$ a `hard decision'  made for the $i$th data symbol under the hypothesis that the channel phase is $\theta$.  Note that $\hat{d}_i(\theta)$ does not depend on $\rho$.  As defined, $\hat{u}_i(\theta)$ is not strictly inside the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$, but this is not of consequence, as we intend its value to be considered equivalent modulo $2\pi$.  With this in mind,
\[
\hat{u}_i(\theta) = \round{\angle{y_i} - \theta }
\]
which is equivalent to the definition from~\eqref{eq:hatdfinxtheta} modulo $2\pi$.

@@

Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the sample MSE of the non-data-aided phase estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  This estimator requires selection of a function $F$ that transforms the amplitude of each sample prior to the final estimation step.  % (see equation (2) in~\cite{ViterbiViterbi_phase_est_1983}).  
Viterbi and Viterbi propose several viable alternatives, from which we have chosen $F(x) = 1$.  The Viterbi and Viterbi estimator is only applicable in the non-data-aided setting, i.e. when $\abs{P} = 0$.  The sample MSE of the least squares estimator (when $\abs{P} = 0$) and the Viterbi and Viterbi estimator is similar.  The least squares estimator appears slightly more accurate for some values of SNR.
Figures~\ref{fig:plotphaseBPSK},~\ref{fig:plotphaseQPSK}~and~\ref{fig:plotphase8PSK} also display the non-data aided Cram\'{e}r-Rao lower bound (CRB) for carrier phase given in~\cite{Cowley_crbs_phase_freq_1996}.  These CRBs are derived for the non-data aided setting and are only valid for estimators that do not make use of pilots symbols, i.e., $\abs{P}=0$.

@@

% The function below is derived by Rice, Cowley, Moran and Rice~\cite{950345}.  It is involved in the computation of CRBs for 8PSK (and QAM).  The function $\Lambda : \complex \mapsto \reals$ is defined as
% \[
% \Lambda(x) = -\frac{A(x)^2}{\sigma^2D(x)^2} + \frac{B(x)}{\sigma^2 D(x)} - \frac{C(x)}{D(x)}
% \]
% where
% \[
% A(x) = \sum_{a\in Q}\big( s(ax)\Im(ax) + s(a^*x)\Im(a^*x) \big)
% \]
% \[
% B(x) = \sum_{a\in Q}\big( c(ax)\Im(ax)^2 + c(a^*x)\Im(a^*x)^2 \big)
% \]
% \[
% C(x) = \sum_{a\in Q}\big( s(ax)\Re(ax) + s(a^*x)\Re(a^*x) \big)
% \]
% \[
% D(x) = \sum_{a\in Q}\big( c(ax) + c(a^*x) \big)
% \]
% and where
% \[
% s(x) = \sinh(\Re{x}/\sigma^2)
% \]
% and 
% \[
% c(x) = \cosh(\Re{x}/\sigma^2)
% \]
